---
title: "Interactive Graphics with ggplot2+gridSVG"
author: "Michael Sachs"
date: "2 July 2015"
output: 
    ioslides_presentation:
      css: styles.css
      logo: US-NIH-NCI-Logo.svg
      self_contained: no
      widescreen: yes
---


```{r init, include = FALSE}
library(ggplot2)
library(gridSVG)
library(grid)
library(survival)
library(dplyr)

```
<script src="d3.min.js"></script>

# The idea

## Steps to Interactive graphics

1. Raw data
    ```{r, echo = FALSE}
    mtcars %>% select(mpg, hp) %>% head(1)
    ```
2. Statistical transformation/analysis: `{x: .5, y: .5}`
3. Rendering
    ```{r, echo = FALSE, fig.width = 1.5, fig.height = 1.5, fig.align = "center"}
    mtcars %>% select(mpg, hp) %>% head(1) %>% ggplot(aes(mpg, hp)) + geom_point(size = 5) + 
      scale_x_continuous("", labels = NULL) + scale_y_continuous("", labels = NULL) + theme_minimal()
    ```
4. Modify rendered elements in browser

<div class="centered">
```{r, echo = FALSE, fig.width = 2, fig.height = 2, fig.keep='none', results = 'asis', fig.align = "center"}
pexsm <- mtcars %>% select(mpg, hp) %>% head(1) %>% ggplot(aes(mpg, hp)) + geom_point(size = 5) + 
    scale_x_continuous("", labels = NULL) + scale_y_continuous("", labels = NULL) + theme_minimal()
pexsm
grid::grid.force()
gridSVG::grid.export(NULL)$svg
```

</div>

<script src="intro.js"></script>


## Approaches to Interactive Graphs

<div>
<svg height = "50" width = "960">
<text x=100 y=40 class="text1" id="first"> 
htmlwidgets, ggvis, qtlcharts, animint, rCharts, plotly
</text>
<text x=350 y=40 class = "text2" id="second" fill = "red">
ggplot2+gridSVG
</text>
<text x=375 y=40 class = "text2" id="third">
shiny
</text>
</svg>
</div>

<script src="slidersummary.js"></script>

# The process

## ggplot2 to svg

```{r ggtosvg, fig.keep='none', results = 'asis', fig.width=7.5, fig.height=3.5}
ggplot(mtcars, aes(mpg, hp)) + geom_point(size = 5)
grid::grid.force()
pts <- grid::grid.ls(print = FALSE)$name %>%
  grep("geom_point.points", ., value = TRUE)
gridSVG::grid.garnish(pts, cyl = paste(mtcars$cyl), group = FALSE)
gridSVG::grid.export(NULL)$svg
```

Not a `.png`! Generates svg code rendered by the browser


<script>
var points = d3.selectAll("[id^='geom_point.points.100.1.']")
points.on("click", function(){
  // do something 
  alert("cyl: " + d3.select(this).attr("cyl"));
})
</script>

## Manipulate the svg with `d3.js`

- [`d3.js`](https://d3js.org) concepts:
    + Select visual elements
    + Get/modify attributes
    + smooth transitions
    + ... in response to events/data

```js
var points = d3.selectAll("[id^='geom_point.points.100.1.']")
points.on("click", function(){
  // do something 
  alert("cyl: " + d3.select(this).attr("cyl"));
})
```

### Click a point on the previous page to display the number of cylinders


## Changing attributes, transitions

```{r test-a, echo = FALSE, fig.keep='none', results = 'asis', fig.width=8, fig.height=6}
p <- ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point(size = 5) + 
  scale_y_continuous(limits = c(-50, 500)) + scale_x_continuous(limits = c(0, 50))
p

grid.force()
## grid.ls()  find the name of the points grob
grid.export(NULL, prefix = "cars")$svg
```

<button id="moveme">Move the points!</button>
  
  <script>
  
  function randCol() {
    
    return "rgb(" + Math.round(Math.random()*255) + ", " + Math.round(Math.random()*255) + ", " + Math.round(Math.random()*255) + ")";
    
  } 

d3.select("#moveme").on("click", function(){
  
  d3.selectAll("[id^='carsgeom_point.points.160.1.']")
  .transition().duration(500)
  .attr("fill", randCol())
  .attr("y", function(d){ return Number(d3.select(this).attr("y")) + 50 * (Math.random() - .5); })
  .attr("x", function(d){ return Number(d3.select(this).attr("x")) + 25 * (Math.random() - .5); });
  
});
</script>

# Useful applications

## Kaplan-Meier Curves {.flexbox .vcenter}

```{r test-km, fig.keep='none', results = 'asis', warning = FALSE, message = FALSE, echo = FALSE, fig.width=10, fig.height=6}
gdat <- survival::lung
fit <- survfit(Surv(time, status) ~ factor(sex), data = gdat)
tdat <- with(fit, data.frame(n.risk, time, surv, strata = rep(1:2, strata)))
targ.times <- seq(min(tdat$time), max(tdat$time), by = 15)[-c(1, 68)]

interp <- function(dat, t){
  
  if(t %in% dat$time){
    
    return(dat[dat$time == t, ])
    
  } else {
    
    t2 <- max(dat$time[dat$time < t])
    
    d2 <- dat[dat$time == t2, ]
    d2$time <- t
    return(d2)
    
  }
  
}

tdat2 <- NULL
for(ttt in targ.times){
  
  tdat2 <- rbind(tdat2, tdat %>% group_by(strata) %>% do({
    
    interp(., ttt)
    
  }))
}
tdat <- tdat2[order(tdat2$strata, tdat2$time),]

p1 <- ggplot(gdat, aes(x = time, status = status, color = factor(sex))) + geom_km()

p1 + geom_text(data = tdat, aes(x = time, y = surv, label = n.risk, hjust = .5, vjust = -1,
                                status = NULL, color = NULL), alpha = 0) + 
  geom_point(data = tdat, aes(x = time, y = surv,
                              status = NULL, color = NULL), alpha = 0)

grid.force()
lsg <- grep("geom_point.points.", grid.ls(print = FALSE)$name, value = TRUE)

grid.garnish(lsg, tip = paste(tdat$n.risk), group = FALSE, global = TRUE, grep = TRUE)
grid.export(NULL, prefix = "km")$svg
```

<script src="kmhover.js"></script>
  
## Transitioning a line {.flexbox .vcenter}
  
```{r trans, fig.keep='none', results = 'asis', echo = FALSE, fig.width=8, fig.height=6}
sdat1 <- survfit(Surv(time, status) ~ 1, data = subset(lung, sex == 1))
sdat2 <- survfit(Surv(time, status) ~ 1, data = subset(lung, sex == 2))
## create stepfunction
dostep <- function(x,y) {
  keep <- is.finite(x) & is.finite(y) 
  if (!any(keep)) return()  #all points were infinite or NA
  if (!all(keep)) {
    # these won't plot anyway, so simplify (CI values are often NA)
    x <- x[keep]
    y <- y[keep]
  }
  n <- length(x)
  if (n==1)       list(x=x, y=y)
  else if (n==2)  list(x=x[c(1,2,2)], y=y[c(1,1,2)])
  else {
    # replace verbose horizonal sequences like
    # (1, .2), (1.4, .2), (1.8, .2), (2.3, .2), (2.9, .2), (3, .1)
    # with (1, .2), (.3, .2),(3, .1).  
    #  They are slow, and can smear the looks of the line type.
    temp <- rle(y)$lengths
    drops <- 1 + cumsum(temp[-length(temp)])  # points where the curve drops
    
    #create a step function
    if (n %in% drops) {  #the last point is a drop
      xrep <- c(x[1], rep(x[drops], each=2))
      yrep <- rep(y[c(1,drops)], c(rep(2, length(drops)), 1))
    }
    else {
      xrep <- c(x[1], rep(x[drops], each=2), x[n])
      yrep <- c(rep(y[c(1,drops)], each=2))
    }
    list(x=xrep, y=yrep)
  }
}

step1 <- dostep(sdat1$time, sdat1$surv)
step2 <- dostep(sdat2$time, sdat2$surv)

p1 <- ggplot(lung, aes(x = time, status = status)) + geom_km(se = FALSE) + 
  scale_x_continuous(limits = c(0, 1022)) + 
  scale_y_continuous(limits = c(0, 1))
p1 
grid.force()
#grid.ls()

grid.garnish("GRID.polyline.465",
             tipy1 = paste(round(step1$y, 3), collapse = ","), 
             tipx1 = paste(step1$x, collapse = ","),
             tipy2 = paste(round(step2$y, 3), collapse = ","), 
             tipx2 = paste(step2$x, collapse = ","),
             group = TRUE, global = TRUE, grep = TRUE)
grid.export(NULL, prefix = "km2")$svg
```

<form action="">
<input type="radio" name="sex" value="overall" checked>Overall<br/>
<input type="radio" name="sex" value="male">Male<br/>
<input type="radio" name="sex" value="female">Female<br/>
</form>
  
  
<script src="kmtransition.js"></script>
  

  
## ROC Curves 
  
```{r rocex, eval = FALSE}
library(plotROC)

roc.ci <- calculate_roc(ex.data$M1, ex.data$D, 
                        ci = TRUE, alpha = 0.05)
ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)

cat(
  export_interactive_roc(ci.rocplot, 
                         prefix = "aci", 
                         omit.d3 = TRUE)
)
```

## Click anywhere {.flexbox .vcenter}

```{r int-nob, fig.keep='none', results = 'asis', warning = FALSE, message = FALSE, echo = FALSE, fig.width=8, fig.height=6}
library(plotROC)
set.seed(140)
D.ex <- rbinom(200, 1, .5)

paired.data <- data.frame(M1 = rnorm(200, mean = D.ex), 
                          D = D.ex)

roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
                        ci = TRUE, alpha = 0.05)
ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)

cat(
  export_interactive_roc(ci.rocplot, 
                         prefix = "aci", 
                         omit.d3 = TRUE)
)
```

## Multiple ROC curves {.flexbox .vcenter}

```{r multistart, include = FALSE}
D.ex <- rbinom(100, 1, .5)

paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
                       M2 = rnorm(100, mean = D.ex, sd = .4), 
                       M3 = runif(100), D = D.ex)

estimate.list <- calculate_multi_roc(paired.data, c("M1", "M2", "M3"), "D")
```

```{r multi3, message = FALSE, warning = FALSE, fig.keep='none', results = 'asis', echo = FALSE, fig.width=8.5, fig.height=6}
colorplot <- multi_ggroc(estimate.list, 
                         xlabel = "1 - Specificity", 
                         ylabel = "Sensitivity")
cat(
  export_interactive_roc(colorplot, lty = rep(1, 3), prefix = "multi3",
                         color = c("black", "purple", "orange"), 
                         legend = TRUE, omit.d3 = TRUE, width = 8.5)
  )
```


## Tips

- Bind data to svg elements with `grid.garnish(key = value)`
    + extract with 
    ```js 
    .attr("key")
    ```
- Create hidden graph elements in `R` (`alpha = 0`), 
    + show them interactively 
    ```js 
    .attr("opacity", 1)
    ```
- Height and width attrs of `panel.background.rect` useful for creating d3 scales
- Use `fig.keep='none', results = 'asis'` in `knitr`

  
## Summary

- Use statistical and rendering power of `R` then create SVG for use in browser
    + Visual consistency across print and web media
- Use svg modification power of `d3.js` to provide interactivity
    + Requires writing a *bit* of JS
    
### Path forward
- Create `R` packages to accomplish specific task, e.g. `plotROC`
- Abstract some of these ideas to make something more broadly useful 


## Acknowledgements

This would not be possible without the following:

  - [ggplot2](http://ggplot2.org/), Hadley Wickham
  - [gridSVG](http://sjp.co.nz/projects/gridsvg/), Simon Potter
  - [d3.js](http://d3js.org), Mike Bostock
  
### Contact

- @sachsmc ([github](https://github.com/sachsmc), [twitter](https://twitter.com/sachsmc))
- [michael.sachs@nih.gov](mailto:michael.sachs@nih.gov)
- [Slides](https://sachsmc.github.io/UseR2015-Talk) and [Source](https://github.com/sachsmc/UseR2015-Talk)





